### 2023-02-09 article.tex
Progress

% This section has paragraph headers. This helps me, the author,
% to focus the contents of my paragraphs.
% In the final version of this manuscript, these will be removed.

@@

% Alex and Blake?
Two different researchers in genetic epideomology (more on that field later), 
write two equally good manuscripts
that involve computation.
Both manuscripts are accepted by an equally prestigous journal after peer view. 
One researcher, however, does not supply the
computer code (from now on: 'code') that was used to generate the results,
where the other does.
Are the conclusion of these papers to be trusted equally?
Is this difference relevant and worth the effort?
This chapter discusses the effect of supplying the computer 
code  
and claims that code is paradata and that this paradata 
is just as important as the paper itself.

@@

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, documented in a useful 
way (and shaped by the data) and be friendly to computers \cite{huvila2022improving}.
These suggestions appear to work well for code too
and will be discussed in detail below.

@@

This paper uses genetic epidemiology as a specific example,
to illustrate that the code is paradata,
and that this type of paradata is relevant.
However, any field that uses computation in its experiments
could be used as an example.

@@

Going back to the example scientists in the introduction,
where one does and the other does not publish the code,
it is not common to publish the code of an experiment or analysis 
\cite{stodden2011trust,read2015sizing} (with a pleasant exception 
being \cite{conesa2019making}).
For example, in computer graphics, 
a field intimately familiar with computer code,
only 192 of 454 SIGGRAPH papers supply computer code \cite{bonneel2020code}.
Another study analysed the reproducibility of registered reports
in the field of psychology, 
where still only 37 out of 62 studies supplied the code 
to redo the analysis \cite{obels2020analysis}.
Unpublished code has been the cause of some saddening examples,
such as an algorithm that detects breast cancer from images 
better than a human expert, 
yet failed to ever be reproduced \cite{haibe2020importance}.

@@

Genetic epidemiology is a field within biology that
tries to determine the mechanism and spread of heritable traits.
For example, we know that lactose intolerance in adults is
caused by a decline in the production of lactose-degrading enzymes,
and is most commonly found in south-east asia and south africa \cite{storhaug2017country}.
The trait is caused by the genetic make-up, or 'genotype'
(see Table \ref{tab:definitions} for the definitions of terms used in this paper), 
of a person.
The trait, also called 'phenotype' (again, see Table \ref{tab:definitions}), 
in this example is lactose intolerance at adult age,
yet any human property, such as weight or height can be studied.
When an association between genotype and phenotype is found,
these associations are subsequently used to 
create so-called 'gene panels', 
where the gene causing 
an association is measured specifically, to, among others,
detect people at risk for the associated phenotype.
The rest of this section describes a genetic epidemiology study 
in more detail, with special focus on the computational experiment.

@@

The first type of primary data, the genotypes, 
consists out of single nucleotide polymorphisms (SNPs, pronounced as 'snips').
A SNP has a name and a location on the DNA, at which there is a certain nucleotide.
DNA (organized into chromosomes and present in every (nucleated) cell, 
\iffalse
(see figure \ref{fig:eukakyote_dna}))
\fi
consists out of billions of nucleotides.
There are four types of nucleotides, 
called adenosine, cytosine, guanine and thyrosine, all commonly abreviated
as A, C, G and T respectively.
One SNP example is \verb|rs12133641|, which is a SNP located at position 
154,428,283, where 67 percent of the people within this study have an A,
and 33 percent have a G (also from \cite{ahsan2017relative}, Table S3).

@@

The second type of data, the phenotypes, 
are concentrations of proteins in the blood. 
DNA contains the code for building proteins
\iffalse
(see Figure \ref{fig:dna_to_protein})
\fi
, as well as the rate
at which a protein is created. Some proteins end up in the blood and
their presence can be used to assess the health of an individual.
IL6RA is one such protein and, spoiler alert, its concentration
is associated with the SNP mentioned earlier.

@@

The conclusions drawn from this paper,
may end up in the clinic.
For the sake of having a clear (yet fictitious) example,
let's assume that a high level of IL6RA 
is associated with a disease that develops later in life,
yet is preventable by lifestyle changes.
Would this be the case, we can create a tailored 
experiment, called a gene panel, that specifically measures
SNP \verb|rs12133641|. 
If the gene panel shows an individual has two guanines, 
we know that this person is likelier to develop higher levels of
IL6RA and is likelier to benefit from the lifestyle changes.

From this simple example, it will be easier
to measure the level of IL6RA in the blood, than using a gene panel,
as blood tests are easier and cheaper.
However, there are associations published for many diseases,
in which one (e.g. phenylketonuria) or many (e.g. \cite{bruce2009metabolic}) 
SNPs contribute to being more likely to develope a disease in the future. 
Here, the phenotype (having a disease in the future) 
is impossible to detect at the present
and associations found in earlier studies are used to create a gene panel.
As creating a gene panel is costly, those associations better be correct.

@@

The experiment described above is run by code. 
It was code that detected the relationship between the genotype
(in this case, SNP \verb|rs12133641|) 
and the phenotype (in this case, the concentration of IL6RA).
To be more precise, it was code that read the data,
subsetted the data, removed outliers, 
performed the statistics and generated the plots.
All these steps are described in the paper, 
but the code itself is unpublished
and the paper does not mention any way to obtain it.
The reponse rate of correnspondence authors with a request for data is around 50\% 
\cite{manca2018non, teunis2015corresponding} 
(note that for the field of emergency medicine, 
these rates are 73\% \cite{o2003email}).

@@

A genetic epidemiology works with sensitive data as well:
the genetic sequences of participants are private.
For research to be reproducible one needs both the code and the data
to reproduce the (hopefully) same results.
This problem is discussed in section \ref{sec:sensitive-data}.

@@

Again, paradata should be comprehensive, documented in a useful way
and friendly to computers \cite{huvila2022improving}.
In this section, these ideal properties of paradata are applied to code.
The resulting recommendations are compared with
existing best practices in writing code.

Code as paradata has multiple uses: to reproduce, to replicate and to 
extend \cite{benureau2018re}.

To reproduce an experiment is to obtain the same publish result,
by using the same code and data as used for that publication 
\cite{peng2006reproducible}.

(1) to reproduce an experiment, by running the unchanged code on the same data as
used in a published research,
(2) to replicate an experiment, by running the unchanged code on different data as
used in a published research,
(3) to extend an experiment, by changing the code.

@@

For code to be ideal paradata, it must be usefully documented.
What is useful is determined by its intended use: 
to reproduce, replicate or extend \ref{sec:making-code-useful-paradata}.
The amount of documentation is, however, also determined
by the recommended practices for writing code in general.

\subsubsection{... for reproducing}

For reproducing purposes, the minimal amount of documentation
is determined by software development best practices.
When an independent researcher reproduces an experiment, 
it is possible that no code is read at all.
Due to this, one could argue that no documentation is needed.
However, all code should be documented 'adequately' \cite{peng2006reproducible},
ideally writing code in such a way that it becomes 
self-explanatory \cite{wilson2014best}
and for the remaining code to document the reasons behind it, 
its design and its purpose \cite{wilson2014best}.

\subsubsection{... for replication}

\subsubsection{... for extending}

There exists a clear relationship between the popularity 
and the documentation effort \cite{cosentino2017systematic,hata2015characteristics}.

More than half of all scientists have their code 
rarely or never reviewed \cite{vable2021code},
although code reviews are known to
accelerate learning of the developers,
improve the quality of the code
and resulting to an experiment that is likelier to be reproducible
\cite{vable2021code}.
One notable exception is rOpenSci \cite{ram2018community},
which is an organisation that does peer-review academic code 
and facilitates the optional publication of a paper.
Although both code and paper are peer-reviewed,
most effort is spent on reviewing the code,
as can be seen in, for example, 
the \verb|mcbette| R package \cite{bilderbeek2020mcbette}.
Within software development best-practices, 
code reviews are recommended as well \cite{wilson2014best}

@@

\paragraph{Coding style}

Code is just 'another kind of data' and should be designed 
as such, for example, by using tools to work on it \cite{wilson2022twelve}.
A first example is to use a tool to enforce a coding style 
(e.g. the Tidyverse style guide \cite{wickham2019advanced} for R,
or PEP 8 \cite{van2001pep} for Python).
Following a consistent coding style improves software quality \cite{fang2001}.
A second example is to use a tool to enforce a low cyclomatic complexity.
The cyclomatic complexity is approximately defined 
as the number of independent paths that
the code can be executed. 
For example, code that has only one \verb|if| statement
has a cyclomatic complexity of 2, as the condition within the \verb|if|
statement can be true or false,
resulting in the body of the \verb|if| statement being either
exectuted or ignored.
The cyclomatic complexity correlates with code complexity,
where more complex code is likelier to contain or give rise to bugs 
\cite{abd2018calculating,chen2019empirical,zimmermann2008predicting}

\paragraph{Error handling}

A feature that deserves more attention
is that 'code must act as a teacher for future developers' \cite{sadowski2018modern}.
Error handling is one of the mechanisms to do so.
A simple example is when one (e.g. a BSc student) 
needs to calculate the variance of a distribution.
Commonly one needs to call a function to do so, 
e.g. \verb|var| in R, that needs the values of the distribution
as input.
However, to calculate the variance of a distribution, at least two values
are required (else a division by zero occurs in the calcation of the variance). 
When the variance of (zero or) only one value is requested,
an error message can helpfully indicate that the function 
needs at least two values to work as expected.

\subsubsection{Code must be hosted}

Code should be distributed in standard ways \cite{peng2006reproducible}.
The standard way to publish code is to use a code hosting website.
There are multiple code hosting websites, 
with GitHub being the most popular one \cite{cosentino2017systematic}.
GitHub is a website where users can create dedicated pages (called 'repositories')
to upload code and interact with it.
The use of such code hosting websites
has increased strongly \cite{russell2018large},
accommodates collaboration \cite{perez2016ten}
and improves transparency \cite{gorgolewski2016practical}.
See \cite{cosentino2017systematic} for an extensive overview of
research conducted on GitHub.

@@

\subsubsection{Code must be tested}

Coding errors are extremely common and contribute to the reproducibility crisis in science \cite{vable2021code}

@@

% \paragraph{Code coverage helps assess code quality}

@@

% \paragraph{Metrics help assess the honest division of labour}

@@

@@

>>>>

@@

% Alex and Blake?
Two different researchers in genetic epideomology (more on that field later), 
write two equally good manuscripts
that describe an experiment with computational steps.
Both manuscripts are accepted by an equally prestigous journal after peer view. 
One researcher, however, does not supply the
computer code (from now on: 'code') that was used to generate the results,
where the other does.
Are the conclusion of these papers to be trusted equally?
Is this difference relevant and worth the effort?
This chapter discusses why code, a type of paradata, 
should be supplied
and how it should be supplied,
and claims that code is just as important as the paper itself.

@@

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, documented in a useful way, 
the documention and data should have co-evolved, 
and the paradata should be good to computers \cite{huvila2022improving}.
These suggestions appear to work well for code too
and will be discussed in detail below.

@@

This paper uses genetic epidemiology as a specific example,
to illustrate that the code is paradata,
and that this type of paradata is relevant.
However, any field that uses computation 
on sensitive data in its experiments,
could be used as an example.

@@

Going back to the example scientists in the introduction,
where one does and the other does not publish the code,
it is not common to publish the code of an experiment or analysis 
\cite{stodden2011trust,read2015sizing} (with a pleasant exception 
being \cite{conesa2019making}).
For example, in computer graphics, 
a field intimately familiar with computer code,
192 of 454 SIGGRAPH papers supply computer code \cite{bonneel2020code}.
Another study analysed the reproducibility of registered reports
in the field of psychology, 
where 37 out of 62 studies supplied the code 
to redo the analysis \cite{obels2020analysis}.
Unpublished code has been the cause of some saddening examples,
such as an algorithm that detects breast cancer from images 
better than a human expert, 
yet failed to ever be reproduced \cite{haibe2020importance}.

@@

Genetic epidemiology is a field within biology that
tries to determine the spread of heritable traits 
and their underlying biological mechanism.
For example, we know that lactose intolerance in adults is
caused by a decline in the production of lactose-degrading enzymes,
and is most commonly found in south-east asia and south africa \cite{storhaug2017country}.
The trait is caused by the genetic make-up, or 'genotype'
(see Table \ref{tab:definitions} for the definitions of terms used in this paper), 
of a person.
The trait, also called 'phenotype' (again, see Table \ref{tab:definitions}), 
in this example is lactose intolerance at adult age,
yet any human property, such as weight or height can be studied.
When an association between genotype and phenotype is found,
these associations are subsequently used to 
create so-called 'gene panels', 
where the gene causing 
an association is measured specifically, to, among others,
detect people at risk for the associated phenotype.
The rest of this section describes a genetic epidemiology study 
in more detail, with special focus on the computational experiment.

@@

The first type of primary data, the genotypes, 
consists out of single nucleotide polymorphisms (SNPs, pronounced 'snips').
A SNP has a name and a location on the DNA, at which there is a certain nucleotide.
DNA (organized into chromosomes and present in every (nucleated) cell, 
\iffalse
(see figure \ref{fig:eukakyote_dna}))
\fi
consists out of billions of nucleotides.
There are four types of nucleotides, 
called adenosine, cytosine, guanine and thyrosine, all commonly abreviated
as A, C, G and T respectively.
One SNP example is \verb|rs12133641|, which is a SNP located at position 
154,428,283, where 67 percent of the people within this study have an A,
and 33 percent have a G (also from \cite{ahsan2017relative}, Table S3).

@@

The second type of data, the phenotypes, 
are concentrations of proteins in the blood. 
DNA contains the code for building proteins
\iffalse
(see Figure \ref{fig:dna_to_protein})
\fi
, as well as the rate
at which a protein is created. Some proteins end up in the blood and
their presence can be used to assess the health of an individual.
IL6RA is one such protein and its concentration may (and will, see below)
be associated with the SNP mentioned earlier.

@@

The conclusions drawn from this paper,
may end up in the clinic.
For the sake of having a clear (yet fictitious) example,
let's assume that a high level of IL6RA 
is associated with a disease that develops later in life,
yet is preventable by lifestyle 
changes (see \cite{pope2003will} for an example in Alzheimer's disease).
Would this be the case, we can create a tailored 
experiment, called a gene panel, that specifically measures
SNP \verb|rs12133641|. 
If the gene panel shows an individual has two guanines, 
we know that this person is likelier to develop higher levels of
IL6RA and is likelier to benefit from the lifestyle changes.

From this simple example, it will be easier
to measure the level of IL6RA in the blood, than using a gene panel,
as blood tests are easier and cheaper.
However, there are associations published for many diseases,
in which one SNP (e.g. phenylketonuria) or many SNPs (e.g. \cite{bruce2009metabolic}) 
contribute to being more likely to develope a disease in the future. 
Here, the phenotype (having a disease in the future) 
is impossible to detect at the present
and associations found in earlier studies are used to create a gene panel.
As creating a gene panel is costly, those associations better be correct.

@@

The experiment described above is run by code. 
It was code that detected the relationship between the genotype
(in this case, SNP \verb|rs12133641|) 
and the phenotype (in this case, the concentration of IL6RA).
To be more precise, it was code that read the data,
subsetted the data, removed outliers, 
performed the statistics and generated the plots.
All these steps are described in the paper (\cite{ahsan2017relative}), 
but the code itself is unpublished
and the paper does not mention any way to obtain it.
One could send an email to the correspending,
yet it is known that the reponse rate of correnspondence authors 
with a request for data in related fields is around 50\% 
\cite{manca2018non, teunis2015corresponding} 
(for the field of emergency medicine, these rates are 73\% \cite{o2003email}).

@@

A genetic epidemiologist works with sensitive data as well:
the genetic sequences of participants are private (\cite{clayton2019law}).
For research to be reproducible one needs both the code and the data
to reproduce the (hopefully) same results.
This problem is discussed in section \ref{sec:sensitive-data}.

@@

Again, useful paradata should be comprehensive, 
documented appropriately, the documentation should co-evolve with the data,
and the data should be friendly to computers \cite{huvila2022improving}.
In this section, these ideal properties of paradata are applied to code.
Code as paradata, however, has multiple uses as well,
which are to reproduce, to replicate and to 
extend a computational experiment \cite{benureau2018re}.
In this section, the recommendations how to make paradata useful
is applied to code and compared to existing best practices in software 
development, for the intended use of the code.

@@

For code to be ideal paradata, it must be usefully documented \cite{huvila2022improving}.
What is 'useful' is determined by its intended use: 
to reproduce, replicate or extend \ref{sec:making-code-useful-paradata}.
The amount of documentation is, however, also determined
by the recommended practices for writing code in general.

For purposes of reproducing (same code, same data) or replicating (same code,
different data) a study, the minimal amount of documentation
is determined by software development best practices.
When an independent researcher reproduces or replicates an experiment, 
it is possible (even ideal!) that no code is read at all.
Due to this, one could argue that no documentation is needed.
However, all code should be documented 'adequately' \cite{peng2006reproducible},
ideally writing code in such a way that it becomes 
self-explanatory \cite{wilson2014best}
and for the remaining code to document the reasons behind it, 
its design and its purpose \cite{wilson2014best}.

For purposes of extending a study and its code, 
documentation becomes even more important,
as the code will be read and modified.
The extent of investing time in documenting code is recommended
to be proportional to the intended reuse \cite{pianosi2020successfully}
and there exists a clear relationship between the reuse 
of code and its documentation 
effort \cite{cosentino2017systematic,hata2015characteristics}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Code and documentation must align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For code to be ideal paradata, 
its creation and documentation need to match, 
'not least because they shape each other' \cite{huvila2022improving}.
This quote resonates strongly with the idea of 
literate programming \cite{knuth1984literate},
in which documentation and code is developed hand-in-hand.
rOpenSci, a community that, among others, reviews programming
code (\cite{ram2013ropensci,ram2018community}), is 
one example of where literate programming is enforced:
code must have extensive runnable examples
as part of the documentation \cite{ropensci_2021_6619350}.
Developing code and documentation at the same time 
leads to both better code and documentation \cite{reenskaug1989environment}.
In general, when developing software, 
it is recommended to 
to write documentation while writing software,
as well as to include many examples \cite{lee2018ten}.

@@

\subsubsection{... to reproduce an experiment}

\paragraph{Code should be hosted}

For code to be extensive enough to reproduce an existing experiment,
it should be distributed in standard ways \cite{peng2006reproducible}.
The standard way to publish code is to use a code hosting website.
There are multiple code hosting websites, 
with GitHub being the most popular one \cite{cosentino2017systematic}.
GitHub is a website where users can create dedicated pages (called 'repositories')
to upload code and interact with it.
The use of such code hosting websites
has increased strongly \cite{russell2018large},
accommodates collaboration \cite{perez2016ten}
and improves transparency \cite{gorgolewski2016practical}.
See \cite{cosentino2017systematic} for an extensive overview of
research conducted on GitHub.

@@

\subsubsection{... to replicate an experiment}

\paragraph{Code must have extensive error handling}

Code must be more extensive when it is (intended to be) 
used on different data,
as then 'code must act as a teacher for future developers' \cite{sadowski2018modern}.
Error handling is one of the mechanisms to do so.
A simple example is when one (e.g. a BSc student) 
needs to calculate the variance of a distribution.
Commonly one needs to call a function to do so, 
e.g. \verb|var| in R, that needs the values of the distribution
as input.
However, to calculate the variance of a distribution, at least two values
are required (else a division by zero occurs in the calcation of the variance). 
When the variance of (zero or) only one value is requested,
an error message can helpfully indicate that the function 
needs at least two values to work as expected.

\subsubsection{... to extend an experiment}

In software development, 
changes are common and should be embraced \cite{beck2000extreme}.
When a scientist intends to reuse code by extending its functionality,
first the code will be assessed,
after which the code may be actually extended.

\subsubsubsection{Assessment of code}

\subsubsubsection{Extending code}

Code is just 'another kind of data' and should be designed 
as such, for example, by using tools to work on it \cite{wilson2022twelve}.

\paragraph{Using a tool for coding style}

A first example is to use a tool to enforce a coding style 
(e.g. the Tidyverse style guide \cite{wickham2019advanced} for R,
or PEP 8 \cite{van2001pep} for Python).
Following a consistent coding style improves software quality \cite{fang2001}.

\paragraph{Using a tool for complexity}

A second example is to use a tool to enforce a low cyclomatic complexity.
The cyclomatic complexity is approximately defined 
as the number of independent paths that
the code can be executed. 
For example, code that has only one \verb|if| statement
has a cyclomatic complexity of 2, as the condition within the \verb|if|
statement can be true or false,
resulting in the body of the \verb|if| statement being either
exectuted or ignored.
The cyclomatic complexity correlates with code complexity,
where more complex code is likelier to contain or give rise to bugs 
\cite{abd2018calculating,chen2019empirical,zimmermann2008predicting}

\paragraph{Code must be tested}

Coding errors are extremely common 
and contribute to the reproducibility crisis in science \cite{vable2021code}.

@@

\paragraph{Code coverage helps assess code quality}

@@

@@

\paragraph{Test-driven development}

TDD improves code quality \cite{alkaoud2018quality,janzen2006test}
and writing comment and documentation mixes well \cite{shmerlin2015document}.

More than half of all scientists have their code 
rarely or never reviewed \cite{vable2021code},
although code reviews are known to
accelerate learning of the developers,
improve the quality of the code
and resulting to an experiment that is likelier to be reproducible
\cite{vable2021code}.
Within software development best-practices, 
code reviews are recommended as well \cite{wilson2014best}

@@

Preserving code by uploading it to a code hosting webite
has many advantages.

\paragraph{Metrics help assess the honest division of labour}
